---
title: "Experiment 1 Mouse Hippocampal DEA"
author: 
  - name: "Neil McAdams"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: cosmo
    highlight: tango
    df_print: paged
---

## Setup

```{r Setup, include=TRUE, warning=FALSE, message=FALSE, results='hide'}
#Libraries for analysis
library(edgeR)
library(readr)
library(variancePartition)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(dplyr)
library(RColorBrewer)
library(gridExtra)
library(grid)
library(PCAtools)
library(matrixStats)
library(tidyr)
library(ggplot2)
library(BayesPrism)

#Set Directories
setwd("C:/path/to/working/directory/")
res.dir <- "C:/path/to/working/directory/results/Experiment1/DESeq2/"; if(!dir.exists(res.dir)){dir.create(res.dir, recursive = TRUE)}

```

```{r Load Data and Preprocess, message=FALSE, warning=FALSE}
#Load and Format Data
#Load in data
metaTable <- read.csv("./input/Experiment1_KS1_metadata.csv"); head(metaTable)
countsTab <- read.delim("./input/E1KS1_counts.tsv", header = TRUE, row.names = 1)
countsTab <-countsTab[,c(metaTable$File)]
colnames(countsTab) <- metaTable$Sample.Name

#Make counts integers
countsTab[] <- lapply(countsTab, function(x) if (is.numeric(x)) as.integer(x) else x)

#Remove Y chromosome genes - Experiment did not control for mouse sex so Y must be removed from analysis
Ygenes <- readLines("./input/YChromosome_genes.txt")
nrow(countsTab);countsTab <- countsTab %>% filter(!(rownames(countsTab) %in% Ygenes)); nrow(countsTab) #decrease by ~ 290

#Filter low counts - Using 15 for a slightly more stringent than the default recommendation of DESeq2
drop <- rowSums(countsTab < 15) > 3
workingCounts <- countsTab[!drop, ]

#make metaData factors
for (i in 1:ncol(metaTable)) {metaTable[,i] <- as.factor(metaTable[,i])}

#Relevel 
metaTable$Group <- relevel(as.factor(metaTable$Group), "deltaCre")
metaTable$Litter <- as.factor(metaTable$Litter)

#write model
design <- ~ Litter + Group

#Create DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = workingCounts,
                              colData = metaTable,
                              design = design)
```

```{r Run Model}
#Run DESeq2 model
dds <- DESeq(dds)

#stabilize variance - for sample level visualizations
vsd <- vst(dds, blind=FALSE)
```

## Exploratory Vizualization

```{r Exploratory Visualization: Dispersion}
#Dispersion
#svg(paste(res.dir,"Dispersion.svg", sep = ""),width = 8, height = 7,bg = "white")
plotDispEsts(dds)
#dev.off()
```

```{r Exploratory Visualizations: Distance}
#Distance matrix
colors <- colorRampPalette(brewer.pal(9, "Blues") )(255)
distanceDat<-assay(vsd)
colnames(distanceDat)<- metaTable$Sample.Name
sample_metadata <- data.frame(Group = c("deltaCre", "deltaCre", "deltaCre", "Cre", "Cre", "Cre"),
                              Litter = c("1", "2", "3", "1", "2", "3"),
                              row.names = metaTable$Sample.Name)

ann_colors<-list(Group = c(Cre="#FFD500", deltaCre="#005BBB"),
                 Litter = c("1" = "#FFBE6A", "2" = "#40B0A6", "3" = "#DB4C77"))

#svg(paste(res.dir,"DistanceMap.svg", sep = ""),width = 12, height = 12, bg = "white") 
pheatmap(cor(distanceDat),
         clustering_distance_rows="euclidean",
         clustering_distance_cols="euclidean",
         cellwidth = 30,
         cellheight = 30,
         annotation_col = sample_metadata,
         annotation_row = sample_metadata,
         annotation_names_col = F,
         annotation_names_row = F,
         annotation_colors = ann_colors,
         annotation_legend = T,
         labels_row = colnames(distanceDat),
         labels_col = c("","","","","",""),
         col = colors)
#dev.off()
```

```{r Exploratory Visualizations: PCA}
#PCA
pcaData <- plotPCA(vsd, intgroup=c("Group", "Litter"), returnData=TRUE)
colnames(pcaData)<- c("PC1", "PC2","combined","Group","Litter","Name")
percentVar <- round(100 * attr(pcaData, "percentVar"))

#svg(paste(res.dir,"PCA.svg", sep = ""),width = 8, height = 7, bg = "white") 
ggplot(pcaData, aes(PC1, PC2, color=Group, shape=Litter)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) +
  coord_fixed()+
  scale_color_manual(values = c(Cre="#FFD500", deltaCre="#005BBB"))+
  theme_bw()
#dev.off()
```

```{r Exploratory Visualizations: PCAtools, warning=FALSE,message=FALSE}
rv <- rowVars(assay(vsd))
select <- order(rv, decreasing=TRUE)[1:500]
p<- pca(assay(vsd)[select,], metadata=colData(vsd), removeVar = 0)

#svg(paste(res.dir,"PCA_scree.svg", sep = ""),width = 8, height = 7, bg = "white") 
screeplot(p)
#dev.off()

#svg(paste(res.dir,"PCA_biplot_1_2.svg", sep = ""),width = 8, height = 11, bg = "white") 
biplot(p, x = "PC1", y = "PC2", colby = 'Group', shape = "Litter", showLoadings = TRUE,labSize = 2, pointSize = 2, sizeLoadingsNames = 2)
#dev.off()

#svg(paste(res.dir,"PCA_pairsplot.svg", sep = ""),width = 16, height = 16, bg = "white") 
pairsplot(p, colby = 'Group', shape = 'Litter', lab = p$yvars, labSize = 3, pointSize = 2)
#dev.off()

#svg(paste(res.dir,"PCA_loadings.svg", sep = ""),width = 8, height = 11, bg = "white") 
plotloadings(p, labSize = 3)
#dev.off()
```

## Call Result Contrasts

```{r Results}
#Results and result plotting
#Generate Results
results <- results(dds, name= "Group_Cre_vs_deltaCre")

#Filter Results - I opt for p-adj <= 0.05 and LFC of +/- 0.5 per literature values of LFC in neuron cell culture
res.nona <- na.omit(results)
res.sig = res.nona[res.nona$padj <= 0.05,] ; nrow(res.sig) # for number of DE Genes at p <= 0.05
res.sig.lfc = res.sig[abs(res.sig$log2FoldChange) > 0.5,]; nrow(res.sig.lfc) # for number of DE Genes at p <= 0.05 & FC =/- 1.5

#write.csv(res.nona, paste(res.dir,"Results.csv", sep = ""), quote=FALSE)
#write.csv(res.sig, paste(res.dir,"sigResults.csv", sep = ""), quote=FALSE)
#write.csv(res.sig.lfc, paste(res.dir,"sig_fcResults_abs05.csv", sep = ""), quote=FALSE)
```

## Result Plotting

```{r Results Plotting: Volcano}
#Volcano plot
source("./fancy_volcano_plot.R") #Custom Volcano plot script prepared by the Patzke lab - I modified it slightly to allow for custom x axis limits and custom titles

#svg(paste(res.dir,"Volcano.svg", sep = ""), width = 8, height = 7,bg = "white") 
fancy_volcano_plot(res.nona, 0.05, 0.5, 3, "DEGs in Cre vs deltaCre +/-0.5")
#dev.off()
```

```{r Results Plotting: Heatmap}
#Heatmap
#data prep
sig_genes <- row.names(res.sig.lfc)
log_sig_counts <- assay(vsd)[sig_genes, ]
colnames(log_sig_counts)<-colData(vsd)$Sample.Name

#svg(paste(res.dir,"DE_Heatmapabs05.svg", sep = ""), width = 8, height = 7, bg = "white") 
pheatmap(na.omit(log_sig_counts),
         cluster_rows=TRUE, cluster_cols=TRUE,   # Clustering
         annotation_col = sample_metadata,
         annotation_names_col = F,
         annotation_colors = ann_colors,
         scale="row",     # Scale by row (gene) to highlight relative expression differences
         show_rownames=F, # Optional: hide gene names for large datasets
         show_colnames = TRUE,
         main="Heatmap of Significant DE Genes")
#dev.off()
```

```{r LFC Shrink}
#LFC Shrinkage - for more accurate GSEA results
results_ashr_shrink <- na.omit(lfcShrink(dds, coef = "Group_Cre_vs_deltaCre"), type = "ashr")
#write.csv(results_ashr_shrink, paste(res.dir,"Results_ashr_shrink.csv", sep = ""), quote=FALSE)

#Ma plot
#svg(paste(res.dir,"MAplot_shrunkenLFCs.svg", sep = ""), width = 8, height = 7, bg = "white")
plotMA(results_ashr_shrink, ylim = c(-3, 3), main = "MA Plot Cre vs DeltaCre")
#dev.off()
```

```{r Custom bar plot genes}
#plotting lfcs
data<- read.csv("./results/Experiment1/DESeq2/sig_fcResults_abs05.csv")
colnames(data)<-gsub("X", "gene",colnames(data))

genes.Neuroinflammation <- c("Anxa1","Atf3","C3ar1","C4b","Ccl12","Ccl9","Cd36","Egr1","F830016B08Rik","Fcgr3","Ifi204","Ifi211","Iigp1c","Il7","Mmp12","Msr1","Oas1a","Oas2","Slc15a3","Socs3","Spp1","Tgfb1","Tnfrsf1b", "Casp1", "Ccl6", "St8sia1", "Slfn5", "Fxyd1", "Gadl1")
genes.ecm<-c("Fn1","Col8a1","Col5a2","Sparcl1","Ltbp2","Thbs1","Ccn2", "Timp3")
genes.Synapse<- c("Adora2a","Atp1a2","Slc7a11","Slc6a1","Slc6a7","Trpm2","Fgf11","Pax2","Shh","Fst","Igf1","Sema3f","Sema3b","Sema4g", "Fxyd7")
```

## Comparative plots

From this point onward it is assumed that the results files for both experiment 1 and experiment 2 have been generated.

```{r comparative barplots}
#read in results
E1KS1_res <- read.csv("./results/Experiment1/DESeq2/Results.csv")
E2KS1_res <- read.csv("./results/Experiment2/KS1/DESeq2/Results.csv")


#filter for genes of interest and prepare for plotting
rE1 <- E1KS1_res %>%
  select(X, log2FoldChange, lfcSE, padj) %>%
  rename(
    log2FC_E1 = log2FoldChange,
    lfcSE_E1  = lfcSE,
    padj_E1   = padj
  ) %>%
  mutate(DE_E1 = !is.na(padj_E1) & padj_E1 < 0.05 & abs(log2FC_E1) >= 0.5)

rE2 <- E2KS1_res %>%
  select(X, log2FoldChange, lfcSE, padj) %>%
  rename(
    log2FC_E2 = log2FoldChange,
    lfcSE_E2  = lfcSE,
    padj_E2   = padj
  ) %>%
  mutate(DE_E2 = !is.na(padj_E2) & padj_E2 < 0.05 & abs(log2FC_E2) >= 0.5)

#filter for each subset of select genes - swap out genes.synapse/genes.ecm/genes.Neuroinflammation for each plot
plot_df <- inner_join(rE1, rE2, by = "X") %>%
  filter(X %in% genes.Synapse) %>%
  mutate(
    ymin_E1 = log2FC_E1 - lfcSE_E1,
    ymax_E1 = log2FC_E1 + lfcSE_E1,
    ymin_E2 = log2FC_E2 - lfcSE_E2,
    ymax_E2 = log2FC_E2 + lfcSE_E2
  )

#transform for plotting
plot_long <- plot_df %>%
  select(
    X,
    log2FC_E1, log2FC_E2,
    ymin_E1, ymax_E1, ymin_E2, ymax_E2,
    padj_E1, padj_E2,
    DE_E1, DE_E2
  ) %>%
  pivot_longer(
    cols = c(log2FC_E1, log2FC_E2),
    names_to = "experiment",
    values_to = "log2FC"
  ) %>%
  mutate(
    ymin = ifelse(experiment == "log2FC_E1", ymin_E1, ymin_E2),
    ymax = ifelse(experiment == "log2FC_E1", ymax_E1, ymax_E2),
    padj = ifelse(experiment == "log2FC_E1", padj_E1, padj_E2),
    DE   = ifelse(experiment == "log2FC_E1", DE_E1, DE_E2),
    log10padj = -log10(padj),
    experiment = recode(
      experiment,
      log2FC_E1 = "hippocampus",
      log2FC_E2 = "glia"
    )
  )

#nLog10 p.values
plot_long$log10padj[is.infinite(plot_long$log10padj)] <- NA
plot_long$log10padj <- pmin(plot_long$log10padj, 10)

gene_order <- plot_df %>%
  mutate(mean_fc = (log2FC_E1 + log2FC_E2) / 2) %>%
  arrange(mean_fc) %>%
  pull(X)

plot_long$X <- factor(plot_long$X, levels = gene_order)

star_df <- plot_long %>%
  filter(DE) %>%
  mutate(
    star_y = ifelse(
      log2FC >= 0,
      ymax + 0.15,
      ymin - 0.15
    )
  )

#plot figure
p <- ggplot(plot_long, aes(x = X, y = log2FC, fill = log10padj)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.2) +
  geom_text(
    data = star_df,
    aes(x = X, y = star_y, label = "*"),
    inherit.aes = FALSE,
    size = 5
  ) +
  coord_flip() +
  geom_hline(yintercept = 0, colour = "gold", linewidth = 0.8) +
  facet_wrap(~experiment, nrow = 1) +
  theme_classic() +
  labs(
    title = "KS1 associated Synapse Genes in Hippocampus and Glia",
    subtitle = "Error bars = ±SE of Log2FoldChange estimation; * = padj < 0.05 and |log2FC| ≥ 0.5",
    x = "",
    y = "Log2 Fold Change",
    fill = expression(-log[10](padj))
  ) +
  scale_fill_gradient(
    low = "grey85",
    high = "blue",
    na.value = "grey90"
  ) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )

p

#ggsave(  "./results/Experiment1/DESeq2/KS1_Synapse_E1vsE2_comparative.svg",
#  plot = p,
#  width = 10,
#  height = 7
#)
```

## Granulator Plotting

This chunk requires the output from the deconvolution workflow and plots the estimated cell type proportions based on correllating expression from our bulk data with single cell expression data from the mouse cortex and hippocampus.

```{r plot granulator results from deconvolution}
#This workflow was ran on HPC but plotted here for ease of visualization
granOut<-readRDS("./input/my.bp.6828.RDS")

theta_type <- get.fraction(
  bp = granOut,
  which.theta = "final",
  state.or.type = "type"
)

if (nrow(theta_type) < ncol(theta_type)) {
  theta_type <- t(theta_type)
}

theta_df <- as.data.frame(theta_type)
theta_df$Sample <- rownames(theta_df)


# Convert to long format
plot_df <- reshape(
  theta_df,
  varying = colnames(theta_type),
  v.names = "Fraction",
  timevar = "CellType",
  times = colnames(theta_type),
  direction = "long"
)
plot_df$group <- c("WT" , "WT", "MUT", "MUT", "WT", "MUT","WT" , "WT", "MUT", "MUT", "WT", "MUT","WT" , "WT", "MUT", "MUT", "WT", "MUT")

rownames(plot_df) <- NULL

#Stack order
plot_df$CellType <- factor(
  plot_df$CellType,
  levels = c("Non-Neuronal", "Glutamatergic", "GABAergic")
)

plot_df <- plot_df %>%
  arrange(group, Sample) %>%   #WT first, then MUT
  mutate(Sample = factor(Sample, levels = unique(Sample)))

#Plot
p <- ggplot(plot_df, aes(x = Sample, y = Fraction * 100, fill = CellType)) +
  geom_bar(stat = "identity", width = 0.8) +
  labs(
    x = "samples",
    y = "estimated cell type proportions (%)",
    fill = "CellType"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )

p
# Save plot
#ggsave("bayesprism_celltype_stackedbar.png", p, width = 8, height = 6, dpi = 600)

```

